---
title: "Playlivemanager Exploration"
output: html_document
---

### Introduction

This is a data exploration for the Udacity Data Analyst Nano Degree.
Data analyzed stems from the Online Fantasy Football Manager www.playlivemanager.com and contains data on Football (Soccer) players from the German Bundesliga for the first half of the 2015/2016 season.

Data has been gathered in JSON format via the API, i.e. at http://www.playlivemanager.com/api/players/64440/stats/round (where 64440 is the respective player id). The JSON data has been converted to a table and information such as home club, position and player name has been matched via webscraping from www.playlivemanager.com.

This data analysis explores drivers for 'total_earnings', which is the objective to maximize in the fantasy football game.


#### Loading Data, libraries and global settings

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.path='Figs/',
echo=FALSE, warning=FALSE, message=FALSE)

```

```{r Load Data & Libraries}
suppressMessages(library("ggplot2"))
suppressMessages(library("dplyr"))
suppressMessages(library("reshape2"))
plm <- read.csv("plm.csv")
dim(plm)
names(plm)
str(plm)
```

### 1 Analysis and exploration of data


#### First look

Initial investigation of total earnings distribution:
```{r init, echo}
qplot(x = total_earnings, data = plm)
```

It looks like lots of players have around 0 points. Only 11 players can play per match, so a lot of players in the dataset will not play on a particular matchday and therefore earn no points. The 'status' variable could help us identify if that explains the many 0s. Let's zoom in on the values close to zero and break it down by status:

```{r byStatus}
qplot(x = total_earnings, data = plm,
      binwidth = 10000,
      color = I('black'), fill = I('lightblue')) +
  facet_wrap(~status, ncol = 1) +
  scale_x_continuous(limits = c(-10000,10000))
```

Indeed, it looks like players who are 'not_present' or on the 'bench' are almost solely responsible for the many obersvations with 0 earnings. Let's check status for just the 0 values:

``` {r status bargraph}
qplot(x = status,data = subset(plm, plm$total_earnings == 0))
```

This highlights the importance of picking players that are likely to play. In terms of analyzing the effect of other variables, these observations can be confounding, so we will take them out from here on onwards.


Let's look at the plot again excluding players on the bench or not present.
```{r nozeros}
plm.played <- subset(plm, !status %in% c("bench", "not_present"))
qplot(x = total_earnings, data = plm.played)
summary(plm.played$total_earnings)
```

This looks much better. The distribution is roughly normal, slightly skewed to the left and with a long tail to the right. It's interesting to note as well that some players have negative values. This can happen if they get yellow cards, red cards or score own goals, for example.


#### Analysis by Club

If a team scores or wins, all players of that team get points. Since some clubs perform better than others throughout a season, we would expect that players playing for a strong club to have higher earnings than players from a weaker club.

```{r byClubbox}
ggplot(data = subset(plm.played, !is.na(Club)),
       aes(x=Club, y=total_earnings)) +
    geom_boxplot() +
    guides(fill=FALSE)
```

It's clear that there are differences by club in this graph. Especially players from FCB and BVB seem to have a far higher median than the other clubs. Let's order the Clubs by median and look at the trendline as well as individual observations.

```{r byClubLine2}
# get clubs in order of median
ClubbyMedian <- with(subset(plm.played, !is.na(Club)), reorder(Club, total_earnings, median))
ggplot(aes(x = ClubbyMedian, y = total_earnings), 
       data = subset(plm.played, !is.na(Club))) +
  geom_point(position = position_jitter(h=0),
             alpha = 0.3, color = "black") +
  geom_line(aes(group = 1), stat = "summary", fun.y = mean, 
            color = "blue", size = 2) +
  geom_line(aes(group = 1), stat = "summary", 
            fun.y = quantile, fun.args=list(probs=0.5), 
            color = "red", size = 1, linetype = 2) +
  geom_smooth(aes(group = 1), method = "lm", size = 2) +
  ylab("Earnings") +
  coord_cartesian(ylim = c(quantile(plm.played$total_earnings, 0.05),
                           quantile(plm.played$total_earnings, 0.95)))
```

Sorting by median makes the trend clear that there are differences by club.

By the same line of reasoning, players may be scoring higher when playing against weaker clubs. We can create a new variable 'opponent' and investigate whether there are differences in total_earnings.

```{r byOpponentbox}
# Get opponent from home/away_shortname
plm <- transform(plm, opponent = ifelse(
  Club == home_shortname, levels(away_shortname)[away_shortname],
  levels(home_shortname)[home_shortname]))
plm.played <- subset(plm, !status %in% c("bench", "not_present"))

ggplot(data = subset(plm.played, !is.na(Club)),
       aes(x=opponent, y=total_earnings)) +
    geom_boxplot() +
    guides(fill=FALSE)
```

We see almost a mirror image of the previous graph, with earnings against BVB and FCB being the lowest. 

Let's look if it's the same trend if we plot it sorted:
``` {r trendopponent}
ClubbyMedian <- with(subset(plm.played, !is.na(Club)), reorder(opponent, total_earnings, median))

ggplot(aes(x = ClubbyMedian, y = total_earnings), 
       data = subset(plm.played, !is.na(Club))) +
  geom_point(position = position_jitter(h=0),
             alpha = 0.3, color = "black") +
  geom_line(aes(group = 1), stat = "summary", fun.y = mean, 
            color = "blue", size = 2) +
  geom_line(aes(group = 1), stat = "summary", 
            fun.y = quantile, fun.args=list(probs=0.5), 
            color = "red", size = 1, linetype = 2) +
  geom_smooth(aes(group = 1), method = "lm", size = 2) +
  ylab("Earnings") +
  coord_cartesian(ylim = c(quantile(plm.played$total_earnings, 0.05),
                           quantile(plm.played$total_earnings, 0.95)))
```

It's not exactly the same order as before, but it's a similar trend. Most points have been earned against teams VFB, TSG, SGE, H96 and BRE. These are indeed the last 5 teams in the current Bundesliga table.

#### Analysis by Position

Players play in different positions and one might expect that their scope for earning points depends on their positions, i.e. a striker (ATT) is more likely to score than a goalkeeper (GOA), while only the latter gets points for a clean sheet.

```{r byPos}
qplot(x = total_earnings, data = plm.played,
      binwidth = 10000,
      color = I('black'), fill = I('#F79420')) +
  facet_wrap(~Pos, ncol = 1)
```

Visually it looks like defenders might have the highest median as there are lots of occurences in the higher score ranges towards the right. However, it's hard to compare as the groups are of different sizes. I.e., there's only 1 GOAlkeeper per match, but several players will play in DEFense and MIDfield, and typically between 1 and 3 in ATTack positions. The histogram confirms this idea:

```{r PosHist}
qplot(x = Pos, data = plm.played)
```

Given this difference in distribution by position, let's look at the medians and compare the range with boxplots. Let's also ignore the NAs:

```{r byPosbox}
qplot(y = total_earnings, x = Pos,
      data = subset(plm.played, !is.na(Pos),
      binwidth = 10000),
      color = I('black'), fill = I('#F79420'),
      geom = 'boxplot')
```

The differences look small, but it seems indeed as if defenders and goalkeepers have higher median scores. However, the ranges are much bigger for midfielders and especially attackers. This would suggest that it may be possible to earn more points with them if selecting the right players. Let's look at actual means and medians:

```{r byPossummary}
by(plm.played$total_earnings, plm.played$Pos, summary)
```

The data somewhat confirms this suspicion. While median and mean score is lowest for attackers by far, the gap gets closer for the 3rd quartile earnings and they have the highest maximum earnings by far. Let's have a look at just the top15%.

```{r byPos15%}
plm.played %>%
  group_by(Pos) %>%
  summarise(quantile90 = quantile(total_earnings, probs = 0.85)) %>%
  arrange(Pos)
```

We can see that in the top 15% range attackers have the highest earnings.


#### Earnings by Player Value

Players in the game have different values:

```{r byValuehist}
ggplot(data = plm.played,
       aes(x=init_Value)) +
    geom_histogram()

summary(plm.played$init_Value)
```

We can see that player values start at 5 and go up all the way to 14. Most observations are from players in the 6.5 to 7.5 range.

We would expect that more expensive players are more likely to earn a lot of points. Let's look at this relationship, excluding NAs again:

```{r byValuebox}
ggplot(data = subset(plm.played, !is.na(init_Value)),
       aes(x=factor(init_Value), y=total_earnings, 
           fill = factor(ceiling(init_Value)))) +
    geom_boxplot() +
    guides(fill=FALSE)
```

The boxplot supports this idea. The median earnings generally increase with the initial value. The highest score in the game was also achieved by a player at 13.5M. The trend is less clear at the higher end beyond 12.0. Few players are valued so high, so there are far fewer observations in this area.

```{r byValueTable}
plm.te_by_value <- plm.played %>%
  group_by(init_Value) %>%
  summarise(mean_earnings = mean(total_earnings),
            median_earnings = median(total_earnings),
            max_earnings = max(total_earnings),
            n = n()) %>%
  arrange(init_Value)

plm.te_by_value
```

If we just look at average and median earnings, the total range of values is smaller and the trend is more clear:

```{r byValueLine}
ggplot(aes(x = init_Value), data = plm.te_by_value) +
  geom_line(aes(y = mean_earnings), color = "blue") +
  geom_line(aes(y = median_earnings), color = "red") +
  ylab("Earnings")
```

The trend is clearly upwards generally, although the variance of median and average is higher at the top end, due to fewer observations leading to less accurate means/medians.

#### Exploring Position and Value

We saw that attackers had the lowest median earnings, but also the highest upside potential for earnings. We also saw that there was a correlation between player value and earnings. So let's look at values for different positions:

```{r valPos}
ggplot(subset(plm.played, !is.na(Pos)),
       aes(x = init_Value)) +
  geom_histogram() +
  facet_wrap(~Pos, ncol = 1, scales = "free_y")
```

Given the different number of players per position we allow the y-scales to adjust to the total number. This makes the histograms more comparable. We can see that the further up a player on the pitch, the more players have higher values. That is, the most expensive players play as attackers and midfielders, and some defenders are still more expensive than the most expensive goalkeeper. This further supports the notion that the best attackers may be worth picking.

#### Value for money

We've established a relationship between Value and Earnings for players. Since there is a constrained budget in picking the fantasy football team, we should also investigate which players are good value for money.

We'll first aggregate our dataset to the player-level. We will include all matches here, even the ones where players did not play. We will exclude players for which we don't have Values.

```{r aggregate}
plm.player <- plm %>%
  filter(!is.na(init_Value)) %>%
  group_by(id, Name) %>%
  summarise(Value = mean(init_Value),
            Total_Earnings = sum(total_earnings),
            mean_earnings = mean(total_earnings),
            Pos = first(Pos)) %>%
  ungroup() %>%
  arrange(id,Name)

head(plm.player)
```

Now let's look at the distribution of players in terms of Value for Money

```{r vfmhist}
ggplot(plm.player, aes(x = Total_Earnings / Value)) +
  geom_histogram(binwidth = 10000)
```

We see that we have a somewhat bell shaped distribution with the majority of values between 0 and ~200,000. There is also a long tail of observations between 200,000 and 300,000. These players can be considered bargains or value for money. Let's zoom in on them and see if we can find a pattern for them:

```{r vfmbox}
plm.vfm <- subset(plm.player, Total_Earnings / Value > 200000)
ggplot(plm.vfm, aes(x = Value, y = Total_Earnings / Value)) +
  geom_boxplot(aes(x = factor(Value)))
```

It is hard to really make out a trend, since there are more players in the lower value brackets and their value for money ratings are more dispersed. It's interesting to see that the highest ratio comes from a player that was worth only 5.5, but the values are all over the place and it's hard to discern a trend.

Let's also check value for money by position:

```{r vfmPos}
ggplot(plm.vfm, aes(x = Pos, y = Total_Earnings / Value)) +
  geom_boxplot(aes(x = factor(Pos)))
```

An interesting trend emerges looking at value for money by position. It looks like the further up the field (ATT, MID), the higher the median value for money. 


#### Explore individual player earnings

Looking even closer at individual players, 17 matches have been played, so for each player we can have up to 17 observations.

```{r byPlayer}
# Set the seed for reproducible results
set.seed(4230)
sample.ids <- sample(plm.played$id, 10)


ggplot(subset(plm, id %in% sample.ids),
       aes(x=matchday, y=total_earnings)) +
  facet_wrap(~Name) +
  geom_line() 
```

We can take away two things from this graph:

1. Many players only played a subset of the matches (i.e. Galvez, Gulselam, Johannsson and Toljan). 
2. There is quite some fluctuation in the earnings of players per matchday.
Let's have a look first at how many matches a player played:

```{r playerobservations}
plm.matchcount <- plm.played %>%
  group_by(id) %>%
  summarize(n = n()) %>%
  arrange(id)

qplot(x=factor(n), data = plm.matchcount)
```

It's interesting to see that the distribution is generally skewed to the right with a significant number of players having played in all 17 matches and lots of players with 13 matches or more. This is in line with the idea of being a "regular starter" and should help in predicting the probability of a player playing in a particular match. It's also interesting to see that a relatively high number of players played only 1  match. Possibly these are fringe players given a chance in a match, but they failed to convince the coach.

Let's now look at what could explain the variations in earnings for when players actually played. Players earn a significant number of points for Clean Sheets, Assists or scoring a goal. Let's look at those variables a bit closer.

```{r Goals/Assists/CleanSheets}
# Replace NAs by 0s for Goals / Assists / Clean Sheets
plm$assist[is.na(plm.played$assist)] <- 0
plm$goal[is.na(plm$goal)] <- 0
plm$clean_sheet[is.na(plm$clean_sheet)] <- 0

# Create new variable to sum up possible achievements
plm <- transform(plm, sum.achievements = assist + goal + clean_sheet)

# Plot histogram of achievements
qplot(x=factor(sum.achievements), data = plm)
```

We can see that most players have 0 achievements on a given match. This is hardly surprising given that we already saw that most players won't even play. From there on, it looks like the scale is decreasing exponentially. Let's have a look by using a log scale:
``` {r histAchivementsSqrd}
qplot(x=factor(sum.achievements), data = plm) +
  scale_y_log10()
```

It looks indeed like there's an exponential decrease in the number of achievements a player can get in an individual match. There are also a high number of NAs, these come from players that have never played a match and we'll ignore them going forward.

Turning back to our sample, let's have a look in how far achievements might explain variations in player earnings:

``` {r sampleAchievements}
# Plot graph as before, but add circles indicating achievements
ggplot(subset(plm, id %in% sample.ids),
       aes(x=matchday, y=total_earnings)) +
  facet_wrap(~Name) +
  geom_line() +
  geom_point(aes(size = sum.achievements), pch = 1)
```

The sample supports this idea. The higher peaks are generally accompanied by a bigger point, except for Galvez and Gulselam. It's notable that these 2 players generally scored quite low, which might be due to the fact that they played less minutes on the field and so had less time to earn points. Let's first look at the distribution of game_play_duration.


``` {r DurationHist}
qplot(x=game_play_duration, data=subset(plm, id %in% sample.ids))
```

Looking at the distribution, it looks strongly U-shaped with bick peaks at 0 and 90 minutes. That is because in a given match there are only 3 substitutions allowed per team, so at least 8 of 11 players will play the full 90 minutes. Similarly, there are typically 8 players on the bench and at least 5 of these 8 players will not get to play at all. In the range in between, we can see that substitutions are usually starting around 60 minutes with the highest value around the 80 minute mark.

Now let's explore that relationship using the game_play_duration variable, using the same visualization as before.

``` {r sampleDuration}
ggplot(subset(plm, id %in% sample.ids),
       aes(x=matchday, y=total_earnings)) +
  facet_wrap(~Name) +
  geom_line() +
  geom_point(aes(size = game_play_duration / 10), pch = 1)

```

Game Play Duration doesn't seem to fully explain the ups and downs. Particularly Galvez' graph still shows a lot of fluctuation even when he played lots of minutes. We saw before that he didn't earn any achievements in any of the matches, so that can't explain the fluctuations either. As we said above, players also earn points for winning a match or when their team scores, so let's add points for wins / draws / losses to the individual achievements. 

Again, let's first look at the distribution:
``` {r histAchievements2}
# Calculate variable for Home or Away match
plm <- transform(plm, home_away = ifelse(Club == home_shortname,
                                           "home", "away"))

# Calculate league points (3 for Win, 1 for Draw, 0 for Loss)
plm <- transform(plm, bl_points = ifelse(home_score == away_score, 1,
                                        ifelse(home_away == "home" &
                                                 home_score > away_score,
                                               3, ifelse(
                                                 home_away == "away" &
                                                   home_score < away_score,
                                                 3, 0))))
qplot(x=factor(sum.achievements + bl_points), data = plm)
```

Comparing this distribution to the previous one it stands out that there are a clearly more observations at the 1 point and the 3 point level. These for the most part should be players that had 0 individual achievements, but got 1 point for a draw or 3 points for a victory as team achievements. The relatively high amount of 4 points can also be explained by players having one individual achievement and their team winning a match. Given that individual achievements are assists, goals or a clean sheet, those increase the likelihood of a match being won.

Using the color scale we will now incorporate these achievements together with playing time into the sample analysis we did before:

``` {r sampleWins}
ggplot(subset(plm, id %in% sample.ids),
       aes(x=matchday, y=total_earnings)) +
  facet_wrap(~Name) +
  geom_line() +
  geom_point(aes(size = game_play_duration, 
                 fill = sum.achievements + bl_points),
             pch = 21) +
  scale_fill_gradientn(colours = colorRampPalette(c("blue", "red"))(5))

```

Adding the team achievement of a win or a draw does add more clarity to the fluctuations. We can see that higher values tend to lie in the reddish area of the colour scale, which includes individual and team achievements. Lower values are associated with the blueish area and / or less playing time. This particularly clears up the fluctuations for Galvez - despite having no individual achievements, he has higher scores when his team won.


### Final Plots and Summary


#### Role of the Club

We first explored the role that a players' club and the opponent he is playing against played in the earnings. We can capture this idea in one graph using tiles:

``` {r MatchUp}
# Calculate matchup matrix
plm.matchup <- plm %>%
  filter(!is.na(Club) & game_play_duration > 0) %>%
  group_by(Club, opponent) %>%
  summarise(median_earnings = median(total_earnings),
            avg_earnings = mean(total_earnings),
            points = first(bl_points)) %>%
  ungroup() %>%
  arrange(Club, opponent)

# Re-level to get clubs in order of total bl_points for plot
plm.sorted <- aggregate(points ~ Club, plm.matchup, sum)

plm.matchup$Club <- factor(plm.matchup$Club,
                      levels = plm.sort[order(plm.sort$points), "Club"])
plm.matchup$opponent <- factor(plm.matchup$opponent,
                      levels = plm.sort[order(plm.sort$points), "Club"])


ggplot(plm.matchup, aes(x = Club, y = opponent, fill = avg_earnings)) +
  geom_tile() +
  scale_fill_distiller(direction = 1) +
  labs(x = "Players' Club", y = "Opponent in a match",
       title = "Average player earnings by match",
       fill = "Average Earnings")
```

In this plot, the teams are listed on the x-axis ordered by their current league table standings (via total points). Opponents are listed on the y-axis in the same order. As the plot is clearly darker towards the bottom right corner and brighter towards the top left, this visualizes the point that the stronger a team and the weaker their opponent, the more points players will earn on average.

This highlights the importance of looking at matchups when selecting players for a team and preferring players from clubs that play a team much lower in the table.


#### Role of Player Value

We already saw a positive trend for player value and earnings. In the follwing visualization we'll add the individual observations as scatterplot and overlay it with a smoother. We'll exclude outliers from the graph to emphasize the relationship. We'll also round the values to full millions to reduce variance at the top: 

```{r byValueLine2}
ggplot(aes(x = ceiling(init_Value), y = total_earnings),
       data = plm.played) +
  geom_point(position = position_jitter(h=0), alpha = 0.5) +
  geom_line(stat = "summary", fun.y = mean, color = "blue", size = 1) +
  geom_smooth(method = "lm") +
  labs(x = "Player Value (Millions)", y = "Player Earnings per match",
       title = "Player Value and median Earnings") +
  coord_cartesian(ylim = c(quantile(plm.played$total_earnings, 0.05),
                           quantile(plm.played$total_earnings, 0.95)))
```

This visualization really drives home two points:

1. Most players are valued 7.5M or less (density of points)
2. The higher a players' value, the higher the median earnings

The smoother line further suggests a correlation. We can run a formal test of correlation:

```{r Value Earnings correlation}
with(plm.played, cor.test(init_Value, total_earnings))
```

The p-value of <2.2e-16 confirms that the true correlation is not equal to 0. We estimate it to be 0.296.


#### Role of individual and club achievements per match

We saw in the individual analysis the role that individual and club-level achievements played in explaining variance in earnings per player. Beyond looking at the small sample, we'll look at all players to see if this holds true.

``` {r extrapolation1}
# Calculate rank variable to sort players by average earnings
plm.player$rank <- rank(-plm.player$mean_earnings)

# Merge with data at match-level
plm.ranked <- merge(plm.player, plm, by = "id")

ggplot(subset(plm.ranked, !is.na(sum.achievements)),
       aes(x=rank, y=total_earnings)) +
  geom_point(aes(color = sum.achievements + bl_points),
                 alpha = 0.8) +
  scale_color_distiller(direction = 1) +
  #  scale_colour_brewer() +  
  geom_line(stat = 'summary', fun.y = mean) +
  scale_y_sqrt() +
  labs(x="Player Rank (by average earnings)", 
       y="Earnings per match (sqrt scale)",
       title="Match achievements and Player Earnings",
       fill="Number of match achievements",
       color = "Average Earnings per match")
```

This visualization really drives home three points:

1. There is significant variance within any individual players earnings (y-scale range per x-value)
2. Match achievements highly correlate with player earnings (more red at upper end of the y-scale)
3. Players with high average earnings play more often (higher density for players with higher ranks)


### Reflection
The data set contains information on 598 Bundesliga players for 17 matches of the currently ongoing season, aggregating to a total of 10,166 observations. As I am playing fantasy football on playlivemanager.com, I was interested in getting a better understanding of players' earnings with the eventual aim to develop a model to predict earnings. 

I started by understanding the distribution of earnings in the data set and immediately found many 0 values. This highlighted that only a subset of players would actually play on a given matchday and highlighted the importance of exploring the relationship of earnings with other variables. Interesting questions and leads opened up as I continued to make observations on plots, such as around Clubs, Positions and Player Value. I explored earnings of players across these variables and got some meaningful insights towards building a strong team in fantasy football: for example, the high amount of players not playing suggests we need to build a model to predict the probability of playing time and include that in a model to predict earnings. The value for money analysis suggested it could be beneficial to play with formations that use more attackers and less defenders. There were also correlations between player values and player clubs and opponents and their earnings.

I then also explored individual players' performance over several matches and was surprised by the huge variance of earnings here. The plots suggest that the variance could be explained by individual achievements, team achievements and playing time to a certain extent. Unfortunately those are hard if not impossible to predict in advance. I therefore expect that a prediction model will only account for some part of the variance. Fortunately there were good indications that the combination of a players' club and its opponent on a given matchday, as well as a players' value correlated with earnings. My aim is to explore this further and then build a linear optimization model to optimize expected earnings subject to buying 11 players at their respective values.