Lesson 5
========================================================

### Multivariate Data
Notes:

***

### Moira Perceived Audience Size Colored by Age
Notes:

***

### Third Qualitative Variable
Notes:

```{r Third Qualitative Variable}
ggplot(aes(x = age, y = friend_count), 
              data = subset(pf, !is.na(gender))) +
  geom_line(aes(color = gender), stat = 'summary', fun.y = median)
```

***

### Plotting Conditional Summaries
Notes:

```{r Plotting Conditional Summaries}
pf.fc_by_age_gender <- pf %>%
  filter(!is.na(gender)) %>%
  group_by(age, gender) %>%
  summarise(mean_friend_count = mean(friend_count),
            median_friend_count = median(friend_count),
            n = n()) %>%
  ungroup() %>%        #not sure why, don't see difference in output..
  arrange(age, gender)

ggplot(pf.fc_by_age_gender,
       aes(x=age, y=median_friend_count, color = gender)) +
         geom_line()
```

***

### Thinking in Ratios
Notes:

***

### Wide and Long Format
Notes:

***

### Reshaping Data
Notes:

```{r}
install.packages('reshape2')
library(reshape2)

pf.fc_by_age_gender.wide <- dcast(pf.fc_by_age_gender,
                                  age ~ gender,
                                  value.var = "median_friend_count")

pf.fc_by_age_gender.wide
```


***

### Ratio Plot
Notes:

```{r Ratio Plot}
ggplot(pf.fc_by_age_gender.wide,
       aes(x=age, y=(female / male))) +
  geom_line() +
  geom_hline(y = 1, linetype = 2, alpha = 0.3)
```

***

### Third Quantitative Variable
Notes:

```{r Third Quantitative Variable}
pf$year_joined = floor(2014 - pf$tenure/365)
```

***

### Cut a Variable
Notes:

```{r Cut a Variable}
summary(pf$year_joined)
table(pf$year_joined)

pf$year_joined.bucket <- cut(pf$year_joined, breaks = c(2004, 2009, 2011, 2012, 2014))
table(pf$year_joined.bucket)
```

***

### Plotting it All Together
Notes:

```{r Plotting it All Together}
table(pf$year_joined.bucket, useNA = 'ifany')

ggplot(aes(x = age, y = friend_count), 
              data = subset(pf, !is.na(year_joined.bucket))) +
  geom_line(aes(color = year_joined.bucket), stat = 'summary', fun.y = median)
```

***

### Plot the Grand Mean
Notes:

```{r Plot the Grand Mean}
ggplot(aes(x = age, y = friend_count), 
              data = subset(pf, !is.na(year_joined.bucket))) +
  geom_line(aes(color = year_joined.bucket), stat = 'summary', fun.y = mean) +
  geom_line(stat = "summary", fun.y = mean, linetype = 2)
```

***

### Friending Rate
Notes:

```{r Friending Rate}
pf$friend_rate <- pf$friend_count / pf$tenure
with(subset(pf, tenure >= 1), summary(friend_rate))

```

***

### Friendships Initiated
Notes:

What is the median friend rate?

What is the maximum friend rate?

```{r Friendships Initiated}


ggplot(aes(x = tenure, y = friendships_initiated / tenure), 
              data = subset(pf, tenure >= 1)) +
  geom_smooth(aes(color = year_joined.bucket))

```

***

### Bias-Variance Tradeoff Revisited
Notes:

```{r Bias-Variance Tradeoff Revisited}

ggplot(aes(x = tenure, y = friendships_initiated / tenure),
       data = subset(pf, tenure >= 1)) +
  geom_line(aes(color = year_joined.bucket),
            stat = 'summary',
            fun.y = mean)

ggplot(aes(x = 7 * round(tenure / 7), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary",
            fun.y = mean)

ggplot(aes(x = 30 * round(tenure / 30), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary",
            fun.y = mean)

ggplot(aes(x = 90 * round(tenure / 90), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary",
            fun.y = mean)

```

***

### Sean's NFL Fan Sentiment Study
Notes:

***

### Introducing the Yogurt Data Set
Notes:

***

### Histograms Revisited
Notes:

```{r Histograms Revisited}
yo <- read.csv('yogurt.csv')
str(yo)
yo$id <- factor(yo$id)

ggplot(yo, aes(x=price)) + geom_histogram(binwidth = 10)
***

### Number of Purchases
Notes:

```{r Number of Purchases}

```

***

### Prices over Time
Notes:

```{r Prices over Time}
yo <- transform(yo, all.purchases = strawberry + blueberry + pina.colada +
                  plain + mixed.berry)


ggplot(yo, aes(x=all.purchases)) + geom_histogram()

ggplot(yo, aes(x=time, y=price)) +
  geom_point(color = 'gray') +
  geom_smooth()
```

***

### Sampling Observations
Notes:

***

### Looking at Samples of Households

```{r Looking at Sample of Households}
# Set the seed for reproducible results
set.seed(4230)
sample.ids <- sample(levels(yo$id), 10)


ggplot(subset(yo, id %in% sample.ids),
       aes(x=time, y=price)) +
  facet_wrap(~id) +
  geom_line() +
  geom_point(aes(size = all.purchases), pch = 1)
```

***

### The Limits of Cross Sectional Data
Notes:

***

### Many Variables
Notes:

***

### Scatterplot Matrix
Notes:

``` {r}
#install.packages("GGally")
library("GGally")

theme_set(theme_minimal(20))

# set seed for reproducible data
set.seed(1836)
pf_subset <- pf[, c(2:15)]
names(pf_subset)
ggpairs(pf_subset[sample.int(nrow(pf_subset),1000),])
```

***

### Even More Variables
Notes:

***

### Heat Maps
Notes:

```{r}
nci <- read.table("nci.tsv")
colnames(nci) <- c(1:64)
```

```{r}
nci.long.samp <- melt(as.matrix(nci[1:200,]))
names(nci.long.samp) <- c("gene", "case", "value")
head(nci.long.samp)

ggplot(aes(y = gene, x = case, fill = value),
  data = nci.long.samp) +
  geom_tile() +
  scale_fill_gradientn(colours = colorRampPalette(c("blue", "red"))(100))
```


***

### Analyzing Three of More Variables
Reflection:
- multiple variables in one plot
- complete correlation matrix & plots  with GGally
- heatmaps for highlighting relationships between variables
- looking more closely at samples (i.e. several observations for one id)
- Variance vs Bias tradeoffs with smoothing (geom_smooth, 30*round(variable/30))
- creating buckets with cut
- adding summaries for groups
- reshaping: one row per case -> one row per combo of variables

***

Click **KnitHTML** to see all of your hard work and to have an html
page of this lesson, your answers, and your notes!

