Lesson 4
========================================================

***

### Scatterplots and Perceived Audience Size
Notes:

***

### Scatterplots
Notes:

```{r Scatterplots}
pf <- read.csv("/home/pi/da-nano/P4-EDA/course/pseudo_facebook.tsv", sep = '\t')
names(pf)
qplot(x = age, y = friend_count, data = pf)
```

***

#### What are some things that you notice right away?
Response:
People in their 20s look more likely to have a lot of friends (>1000).
***

### ggplot Syntax
Notes:

```{r ggplot Syntax}
ggplot(aes(x = age, y = friend_count), data = pf) + geom_point() +
  xlim(13,90)

```

***

### Overplotting
Notes:

```{r Overplotting}
ggplot(aes(x = age, y = friend_count), data = pf) + 
  geom_jitter(alpha = 1/20) +
  xlim(13,90)

```

#### What do you notice in the plot?
Response:
There seems to be a low dip for people in their 40s in terms of friends.
***

### Coord_trans()
Notes:

```{r Coord_trans()}
ggplot(aes(x = age, y = friend_count), data = pf) + 
  geom_point(alpha = 1/20, position = position_jitter(h=0)) +    ## won't work with geom_jitter
  xlim(13,90) +
  coord_trans(y = 'sqrt')
```

#### Look up the documentation for coord_trans() and add a layer to the plot that transforms friend_count using the square root function. Create your plot!

```{r}
ggplot(aes(x = age, y = friend_count), data = pf) + 
  geom_point(alpha = 1/20, position = position_jitter(h=0)) +    ## won't work with geom_jitter
  xlim(13,90) +
  coord_trans(y = 'sqrt')

```

#### What do you notice?

***

### Alpha and Jitter
Notes:

```{r Alpha and Jitter}
ggplot(aes(x = age, y = friendships_initiated), data = pf) + 
  geom_point(alpha = 1/20, position = position_jitter(h=0)) + 
  xlim(13,90) +
  coord_trans(y = 'sqrt')

```

***

### Overplotting and Domain Knowledge
Notes:

***

### Conditional Means
Notes:

```{r Conditional Means}
library("dplyr")

age_groups <- group_by(pf, age)
pf.fc_by_age <- summarise(age_groups,
                          friend_count_mean = mean(friend_count),
                          friend_count_median = median(friend_count),
                          n = n())
pf.fc_by_age <- arrange(pf.fc_by_age, age)

head(pf.fc_by_age)


## Alternative Code

pf.fc_by_age <- pf %>%
  group_by(age) %>%
  summarise(friend_count_mean = mean(friend_count),
            friend_count_median = median(friend_count),
            n = n()) %>%
  arrange(age)

head(pf.fc_by_age)

```

Create your plot!

```{r Conditional Means Plot}
ggplot(aes(x=age, y=friend_count_mean), data=pf.fc_by_age) +
  geom_line()
```

***

### Overlaying Summaries with Raw Data
Notes:

```{r Overlaying Summaries with Raw Data}
ggplot(aes(x = age, y = friendships_initiated), data = pf) + 
  geom_point(alpha = 1/20, position = position_jitter(h=0), color = "orange") + 
  xlim(13,90) +
  coord_trans(y = 'sqrt') +
  geom_line(stat = "summary", fun.y = mean) +
  geom_line(stat = "summary", fun.y = quantile, fun.args=list(probs=0.5),
            color = "blue") +
  geom_line(stat = "summary", fun.y = quantile, fun.args=list(probs=0.1),
            color = "blue", linetype = 2) +
  geom_line(stat = "summary", fun.y = quantile, fun.args=list(probs=0.9),
            color = "blue", linetype = 2)
```

#### What are some of your observations of the plot?
Response:

***

### Moira: Histogram Summary and Scatterplot
See the Instructor Notes of this video to download Moira's paper on perceived audience size and to see the final plot.

Notes:

***

### Correlation
Notes:

```{r Correlation}
cor.test(pf$age, pf$friend_count, "two.sided", "pearson")
with(pf, cor.test(age, friend_count))
```

Look up the documentation for the cor.test function.

What's the correlation between age and friend count? Round to three decimal places.
Response:

***

### Correlation on Subsets
Notes:

```{r Correlation on Subsets}
with(subset(pf, age <= 70), cor.test(age, friend_count))
```

***

### Correlation Methods
Notes:

***

## Create Scatterplots
Notes:

```{r}
ggplot(aes(likes_received, www_likes_received), data = pf) +
  geom_point() +
  coord_trans(x = 'sqrt', y = 'sqrt')
```

***

### Strong Correlations
Notes:

```{r Strong Correlations}
ggplot(aes(likes_received, www_likes_received), data = pf) +
  geom_point() +
  xlim(0, quantile(pf$likes_received, 0.95)) +
  ylim(0, quantile(pf$www_likes_received, 0.95)) +
  geom_smooth(method = 'lm', colour = "red")
```

What's the correlation betwen the two variables? Include the top 5% of values for the variable in the calculation and round to 3 decimal places.

```{r Correlation Calcuation}
with(pf, cor.test(likes_received, www_likes_received))
```

Response:

***

### Moira on Correlation
Notes:

***

### More Caution with Correlation
Notes:

```{r More Caution With Correlation}
#install.packages('alr3')
library(alr3)
```

Create your plot!

```{r Temp vs Month}
data(Mitchell)
head(Mitchell)

ggplot(aes(Month, Temp), data = Mitchell) +
  geom_point()
```

***

### Noisy Scatterplots
a. Take a guess for the correlation coefficient for the scatterplot.
0.1

b. What is the actual correlation of the two variables?
(Round to the thousandths place)

```{r Noisy Scatterplots}
with(Mitchell, cor.test(Month, Temp))
```

***

### Making Sense of Data
Notes:

```{r Making Sense of Data}
ggplot(aes(Month %% 12, Temp), data = Mitchell) +
  geom_point()
```

***

### A New Perspective

What do you notice?
Response:

Watch the solution video and check out the Instructor Notes!
Notes:

***

### Understanding Noise: Age to Age Months
Notes:

```{r Understanding Noise: Age to Age Months}
pf$age_with_month <- pf$age + (1 - pf$dob_month) / 12)
pf$age_with_months <-pf$age + (1 - pf$dob_month / 12)

```

***

### Age with Months Means

```{r Age with Months Means}
pf.fc_by_age_months <- subset(pf) %>%
  group_by(age_with_month) %>%
  summarise(friend_count_mean = mean(friend_count),
            friend_count_median = median(friend_count),
            n = n()) %>%
  arrange(age_with_month)

head(pf.fc_by_age_months)
```

Programming Assignment
```{r Programming Assignment}
ggplot(aes(x = age_with_month, y = friend_count_mean), 
       data = subset(pf.fc_by_age_months, age_with_month < 71)) +
  geom_line()
```

***

### Noise in Conditional Means

```{r Noise in Conditional Means}

```

***

### Smoothing Conditional Means
Notes:

```{r Smoothing Conditional Means}
p1 <- ggplot(aes(x = age_with_month, y = friend_count_mean), 
       data = subset(pf.fc_by_age_months, age_with_month < 71)) +
  geom_line() +
  geom_smooth()

p2 <- ggplot(aes(x = age, y = friend_count_mean), 
       data = subset(pf.fc_by_age, age < 71)) +
  geom_line() +
  geom_smooth()

p3 <- ggplot(aes(x = round(age/5) * 5, y = friend_count), 
       data = subset(pf, age < 71)) +
  geom_line(stat = "summary", fun.y = mean)

library(gridExtra)
grid.arrange(p1,p2, p3, ncol = 1)


```

***

### Which Plot to Choose?
Notes:

***

### Analyzing Two Variables
Reflection:
- ggplot
- scatter plots & how to handle overplotting
- conditional means
- testing for correlation
- using dplyr and chaining commands via %>%
- adding layers, i.e. scatter plot + line for average
- adding means via geom_line(stat = "summary", y.fun = mean)
***

Click **KnitHTML** to see all of your hard work and to have an html
page of this lesson, your answers, and your notes!

