Lesson 3
========================================================

***

### What to Do First?
Notes:
- Load data
***

### Pseudo-Facebook User Data
Notes:

```{r Pseudo-Facebook User Data}
getwd()
list.files()
pf <- read.csv("/home/pi/da-nano/P4-EDA/course/pseudo_facebook.tsv", sep = '\t')
names(pf)
```

***

### Histogram of Users' Birthdays
Notes:

```{r Histogram of Users\' Birthdays}
#install.packages('ggplot2')
library(ggplot2)

qplot(x=dob_day, data = pf) +
  scale_x_discrete(breaks = 1:31)
```

***

#### What are some things that you notice about this histogram?
Response:
Lots on the 1st, which is surprising. Less on the 31st, which is less surprising.
***

### Moira's Investigation
Notes:

***

### Estimating Your Audience Size
Notes:

***

#### Think about a time when you posted a specific message or shared a photo on Facebook. What was it?
Response: Iron Man

#### How many of your friends do you think saw that post?
Response: 300

#### Think about what percent of your friends on Facebook see any posts or comments that you make in a month. What percent do you think that is?
Response: 50%

***

### Perceived Audience Size
Notes:

***
### Faceting
Notes:

```{r Faceting}
qplot(x=dob_day, data = pf) +
  scale_x_discrete(breaks = 1:31) +
  facet_wrap(~dob_month, ncol = 3)
```

#### Let’s take another look at our plot. What stands out to you here?
Response:
So many birthdates on Jan 1 
***

### Be Skeptical - Outliers and Anomalies
Notes:

***

### Moira's Outlier
Notes:
#### Which case do you think applies to Moira’s outlier?
Response:
- bad data about an extreme case (impossible to have 10M visits as max_number of friends is 1M) --> eliminate
***

### Friend Count
Notes:

#### What code would you enter to create a histogram of friend counts?

```{r Friend Count}
qplot(x=friend_count, data=pf)
```

#### How is this plot similar to Moira's first plot?
Response:
- long-tail data; accurate data, but long-tail outliers cause big ranges for histogram buckets
***

### Limiting the Axes
Notes:

```{r Limiting the Axes}
qplot(x=friend_count, data=pf, binwidth = 25) +
  scale_x_continuous(limits = c(0,1000), breaks = seq(0,1000, 50))

```

### Exploring with Bin Width
Notes:

***

### Adjusting the Bin Width
Notes:

### Faceting Friend Count
```{r Faceting Friend Count}
# What code would you add to create a facet the histogram by gender?
# Add it to the code below.
qplot(x = friend_count, data = pf, binwidth = 10) +
  scale_x_continuous(limits = c(0, 1000),
                     breaks = seq(0, 1000, 50)) +
  facet_wrap(~gender, ncol = 2)
```

***

### Omitting NA Values
Notes: Could also just do 'data = na.omit(pf)', but that would exclude any rows with single NA values, which may not be what we want.

```{r Omitting NA Values}
qplot(x = friend_count, data = subset(pf, !is.na(gender)), binwidth = 10) +
  scale_x_continuous(limits = c(0, 1000),
                     breaks = seq(0, 1000, 50)) +
  facet_wrap(~gender, ncol = 2)

```

***

### Statistics 'by' Gender
Notes:

```{r Statistics \'by\' Gender}
table(pf$gender)
by(pf$friend_count, pf$gender, summary)
```

#### Who on average has more friends: men or women?
Response:

#### What's the difference between the median friend count for women and men?
Response:

#### Why would the median be a better measure than the mean?
Response:

***

### Tenure
Notes:

```{r Tenure}
qplot(x=tenure, data = pf,
  color = I('black'), fill = I('#099DD9'))
```

***

#### How would you create a histogram of tenure by year?

```{r Tenure Histogram by Year}
qplot(x=tenure/365, data = pf,
  color = I('black'), fill = I('#F79420'),
  binwidth = 0.25) +
  scale_x_continuous(limits = c(0,7), breaks = seq(0,7,1))
```

***

### Labeling Plots
Notes:

```{r Labeling Plots}
qplot(x=tenure/365, data = pf,
    color = I('black'), fill = I('#F79420'),
    xlab = "Number of Years using FB",
    ylab = "Number of Users in Sample",
    binwidth = 0.25) +
  scale_x_continuous(limits = c(0,7), breaks = seq(0,7,1))
```

***

### User Ages
Notes:

```{r User Ages}
qplot(x=age, data=subset(pf, !is.na(gender)), 
      binwidth = 1,
      color = I('black'), fill = I('#F79420')) +
  facet_wrap(~gender)
```

#### What do you notice?
Response:

***

### The Spread of Memes
Notes:

***

### Lada's Money Bag Meme
Notes:

***

### Transforming Data
Notes:

***

### Add a Scaling Layer
Notes:

```{r Add a Scaling Layer}
#install.packages('gridExtra') 
library(gridExtra)
p1 <- qplot(x = friend_count, data = subset(pf, !is.na(gender)))
p2 <- qplot(x = log10(friend_count), data = subset(pf, !is.na(gender)))
p3 <- qplot(x = sqrt(friend_count), data = subset(pf, !is.na(gender)))

grid.arrange(p1, p2, p3, ncol = 1)

p1 <- ggplot(aes(x = friend_count), data = pf) +
      geom_histogram()
p2 <- p1 + scale_x_log10()
p3 <- p1 + scale_x_sqrt()
```

***


### Frequency Polygons

```{r Frequency Polygons}
qplot(x = friend_count, data = subset(pf, !is.na(gender)), binwidth = 10) +
  scale_x_continuous(limits = c(0, 1000),
                     breaks = seq(0, 1000, 50))

qplot(x = friend_count, y = ..count../sum(..count..),
      data = subset(pf, !is.na(gender)), binwidth = 10,
      geom = 'freqpoly', color = gender) +
  scale_x_continuous(limits = c(0, 1000),
                     breaks = seq(0, 1000, 50))

```

***

### Likes on the Web
Notes:

```{r Likes on the Web}
qplot(x = www_likes, data = subset(pf, !is.na(gender)),
      geom = 'freqpoly', color = gender) +
  scale_x_continuous() +
  scale_x_log10()

table(pf$www_likes, pf$gender)
by(pf$www_likes, pf$gender, sum)

```


***

### Box Plots
Notes:

```{r Box Plots}
qplot(x = gender, y = friend_count,
      data = subset(pf, !is.na(gender)),
      geom = 'boxplot') +
#  scale_y_continuous(limits = c(0,1000)) <- omits values out of scale from median/mean/etc. calculations
  coord_cartesian(ylim = c(0, 1000))
```

#### Adjust the code to focus on users who have friend counts between 0 and 1000.

```{r}

```

***

### Box Plots, Quartiles, and Friendships
Notes:

```{r Box Plots, Quartiles, and Friendships}
qplot(x = gender, y = friend_count,
      data = subset(pf, !is.na(gender)),
      geom = 'boxplot') +
  coord_cartesian(ylim = c(0, 250))
```

#### On average, who initiated more friendships in our sample: men or women?
Response:
#### Write about some ways that you can verify your answer.
Response:
```{r Friend Requests by Gender}
qplot(x = gender, y = friendships_initiated,
      data = subset(pf, !is.na(gender)),
      geom = 'boxplot') +
  coord_cartesian(ylim = c(0, 250))

by(pf$friendships_initiated, pf$gender, mean)
```

Response:

***

### Getting Logical
Notes:

```{r Getting Logical}
summary(pf$mobile_likes)

summary(pf$mobile_likes > 1)

mobile_check_in <- NA
pf$mobile_check_in <- ifelse(pf$mobile_likes > 0, 1, 0)
pf$mobile_check_in <- factor(pf$mobile_check_in)
summary(pf$mobile_check_in)

sum(pf$mobile_check_in == 1) / length(pf$mobile_check_in)

```

Response:

***

### Analyzing One Variable
Reflection:
- qplot for quick histograms (and boxplots)
- by for quick grouping
- facet_wraps(~) for multi-graphs
- geom = 'freqpoly' for frequency by groups (i.e. gender)
- scales with limits & breaks
- using coord_cartesian() rather than x_scale_.. to keep taking all values into account
***

Click **KnitHTML** to see all of your hard work and to have an html
page of this lesson, your answers, and your notes!